---
title: "Practicum I"
output: html_notebook
---

Authors:

Krithika Iyer iyer.kr@northeastern.edu

Nisha Mishra mishra.ni@northeastern.edu

 Load required libraries

```{r}
library(readr)
library(sqldf)
require(RMySQL)
library(ggplot2)

```


 Set up and connect to mysql database


```{r}

# Settings
db_user <- 'root'
db_password <- 'password'
db_name <- 'BirdStrikes'
db_host <- 'localhost' # Workbench Host
db_port <- 3306

# 3. Read data from db
mydb <-  dbConnect(RMySQL::MySQL(), user = db_user, password = db_password,
                 dbname = db_name, host = db_host, port = db_port)

```

// Creating the look up tables airline and airports and load data from csv to mysql table.


1. Creating airline table

```{sql connection=mydb}
CREATE TABLE airline(
aid int PRIMARY KEY NOT NULL,
code VARCHAR(100),
airline VARCHAR(100) NOT NULL
)

```


2. Creating airports table

```{sql connection=mydb}
CREATE TABLE airports(
pid int PRIMARY KEY NOT NULL,
code VARCHAR(100),
name VARCHAR(100) NOT NULL,
city VARCHAR(100) NOT NULL,
state VARCHAR(100) NOT NULL,
country VARCHAR(100) NOT NULL
)
```


3. Extract data from BirdStrikes.csv to dataframe

```{r}

fn = "BirdStrikesData.csv"

df.raw <- read.csv(file = fn,
                   header = T,
                   stringsAsFactors = F)

```

4. Inspect the dataframe and replace the missing values with "unknown"

```{r}
df.raw <- replace(df.raw, df.raw == "","unknown")

head(df.raw, 6)
```

5. Extract airline name from the dataframe and store it in another dataframe

```{r}
n.airlines <- nrow(df.raw)
df.airlines <- data.frame(aid = seq(1,n.airlines),
                         code = "default",
                         airline = df.raw$Aircraft..Airline.Operator
                         )

```

6. Since we are developing an airline lookup table we will drop the duplicate airlines from the dataframe.

```{r}
df.airlines <- df.airlines[!duplicated(df.airlines$airline), ]
rownames(df.airlines) <- NULL
df.airlines
```
DB enable write permission

```{r}
dbExecute(mydb,"SET GLOBAL local_infile = 'ON'")


```



7. Bulk load the data into sql table

```{r}
dbWriteTable(mydb, "airline", df.airlines, overwrite = T)

```

8. Remove the row_names field from the table as it is not required.

```{sql connection=mydb}
ALTER TABLE airline
DROP COLUMN row_names;

```

9. Since the table is overwritten we need to add primary key constraint again.

```{sql connection=mydb}
ALTER TABLE airline
ADD CONSTRAINT PK_airline PRIMARY KEY (aid);
```

10. Add unknown to the look up table to point null values.

```{sql connection=mydb}
INSERT INTO airline (aid, code, airline) VALUES (0,"default", "unknown")


```

11. Extract airport name, state from the dataframe df.raw and store it in another dataframe

```{r}
n.airports <- nrow(df.raw)
df.airports <- data.frame(pid = seq(1,n.airlines),
code = "default",
name = df.raw$Airport..Name,
city = "default",
state = df.raw$Origin.State,
country = "USA"
)
```

12. Remove duplicate airports from the dataframe.

```{r}
df.airports <- df.airports[!duplicated(df.airports$name),]
df.airports
```

13. Bulk load the data into airports sql table

```{r}
dbWriteTable(mydb, "airports", df.airports, overwrite = T)

```

14. Remove the row_names field from the table as it is not required.

```{sql connection=mydb}
ALTER TABLE airports
DROP COLUMN row_names;

```

15.Since the table is overwritten we need to add primary key constraint again.

```{sql connection=mydb}
ALTER TABLE airports
ADD CONSTRAINT PK_airports PRIMARY KEY (pid);

```


16. Add unknown to the look up table to reference null values.

```{sql connection=mydb}
INSERT INTO airports(pid,code,name, city, state, country) VALUES (0,"default", "unknown","default","default","USA")


```

17. Find the Distinct Flight phase in birdstrikes.csv.

```{r}
df.flightStatus <- df.raw[!duplicated(df.raw$When..Phase.of.flight), ]
df.flightStatus$When..Phase.of.flight

```

18. Update the dataframe by categorising the flight phases into 4 categories: take off, inflight, landing and unknown.

```{r}
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == "Climb"] <- "inflight"
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == "Take-off run"] <- "takeOff"
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == "Approach"] <- "inflight"
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == "Descent"] <- "inflight"
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == "Landing Roll"] <- "landing"
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == "Taxi"] <- "unknown"
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == "Parked"] <- "unknown"
df.raw$When..Phase.of.flight[df.raw$When..Phase.of.flight == ""] <- "unknown"

```

19. Creating incident table

```{sql connection=mydb}
CREATE TABLE incident (
iid int PRIMARY KEY NOT NULL,
incidentDate DATE NOT NULL,
depPort bigint NOT NULL,
arrPort bigint NOT NULL,
airline bigint NOT NULL,
aircraft VARCHAR(100) NOT NULL,
flightPhase TEXT NOT NULL,
impact VARCHAR(100) NOT NULL,
FOREIGN KEY(depPort) REFERENCES airports(pid) ON DELETE CASCADE,
FOREIGN KEY(arrPort) REFERENCES airports(pid) ON DELETE CASCADE,
FOREIGN KEY(airline) REFERENCES airline(aid) ON DELETE CASCADE
);

```


20. Extract impact, aircraft, recordid from dataframe and store it another dataframe.

(Pls check if the dataframe has similar coloumn name as mentioned below if the code throws error)

```{r}
n.incidents <- nrow(df.raw)
df.incidents <- data.frame(iid = df.raw$ï..Record.ID,
                           incidentDate = as.Date(df.raw$FlightDate, format = "%m/%d/%Y"),
                           flightPhase = df.raw$When..Phase.of.flight,
                           impact = df.raw$Effect..Impact.to.flight,
                           aircraft = df.raw$Aircraft..Make.Model,
                           depPort = 1,
                           arrPort = 1,
                           airline = 1)
```

21. Link foreign keys and primary keys to respective tables. Here depPort and arrPort will be linked to pid of airport and airline will be linked to aid of airline tabe.

```{r}
for (r in 1:n.incidents) {
  # find the airline PK for that flight in the airlines table
 
  a <- df.airports$pid[which(df.airports$name == df.raw$Airport..Name[r])]
  df.incidents$depPort[r] <- a
  b <- df.airports$pid[which(df.airports$name == df.raw$Airport..Name[r])]
  df.incidents$arrPort[r] <- b
  c <- df.airlines$aid[which(df.airlines$airline == df.raw$Aircraft..Airline.Operator[r])]
  df.incidents$airline[r] <- c
  
}
```


22. Write to database,

```{r}
dbWriteTable(mydb, "incident", df.incidents, overwrite = T)
```

23. Remove the row_names col as it is not required

```{sql connection=mydb}
ALTER TABLE incident DROP COLUMN row_names
```



24. Add primary key constraints.

```{sql connection=mydb}
ALTER TABLE incident 
ADD CONSTRAINT PK_incident PRIMARY KEY (iid);

```

// View the three database.

1. Airline look up table

```{sql connection=mydb}
SELECT * FROM airline

```

2. Airports look up table

```{sql connection=mydb}
SELECT * FROM airports

```

3. Incident table

```{sql connection=mydb}
SELECT * FROM incident 
```


// find the number of bird strike incidents for each airline arriving at LaGuardia airport during any phase of landing.

```{sql connection=mydb}
SELECT airline.airline, COUNT(iid) FROM incident
JOIN airports
ON incident.arrPort = airports.pid
JOIN airline
ON incident.airline = airline.aid
WHERE airports.name LIKE "%LaGuardia%"
AND incident.flightPhase = "landing"
GROUP BY airline.airline

```

// find the airport that had the most bird strike incidents (during any flight phase). Include all commercial airlines, i.e., no business, private, or military flights. 

```{sql connection=mydb}
select airport_name from
(SELECT airports.name as airport_name, COUNT(iid) FROM incident
JOIN airports
ON incident.arrPort = airports.pid
JOIN airline
ON incident.airline = airline.aid
WHERE airline.airline not like '%BUSINESS%'
OR airline.airline not like '%MILITARY%' 
OR airline.airline not like '%PRIVATE%' GROUP BY airports.name order by count(iid) desc LIMIT 1) as a

```

//  find the number of bird strike incidents by year. Include all airlines and all flights

```{sql connection=mydb}
SELECT YEAR(incidentDate), COUNT(iid) from incident group by YEAR(incidentDate);
```



// build a line chart that visualizes the number of bird strikes incidents per year from 2005 to 2011.

```{r}
df <- sqldf(drv="SQLite","SELECT strftime('%Y', incidentDate * 3600 * 24, 'unixepoch') year, COUNT(iid) from `df.incidents` WHERE year >= '2005' AND year <=   '2011'group by year")

df

```
Use the ggplot function of R package ggplot2. 

```{r}

ggplot(data=df, aes(x=df$year, y=df$`COUNT(iid)`, group=1)) +
  geom_line(color="red")+
  geom_point()+
  labs(title="Plot of Number of Birdstrike incidents between 2005 and 2011 (inclusive)",x="Year", y ="Number of Birdstrike Incidents" )

```


// a stored procedure in MySQL that removes a bird strike incident from the database when passed the iid.

```{sql connection=mydb}
CREATE PROCEDURE delete_incident ( in input_id int )
BEGIN
DELETE FROM incident
WHERE iid=input_id;
END

```

The iid which will be deleted

```{sql connection=mydb}
SELECT * FROM incident WHERE iid = 202152

```


Testing the procedure

```{sql connection=mydb}
CALL delete_incident (202152);

```
iid = 202152 deleted.

```{sql connection=mydb}
SELECT * FROM incident WHERE iid = 202152


```


Closing connection

```{r}
dbDisconnect(mydb)

```

